FraCaS: Temporal Analysis

### 2020-03-10 paper2020/paper.tex
texification

\documentclass[a4paper,11pt]{article}
\usepackage[hyperref]{naaclhlt2019}
\usepackage{natbib}
\usepackage{tabularx}

@@

\usepackage[verbose]{newunicodechar}

@@

In montegovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise verb to take two
additional time parameters, one corresponding with the starting time
of the action and one corresponding to its stopping time. For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t0,t1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). We are thus
using a simple newtonian model of time, corresponding to layman
intuition.

@@

The timespan in the context is optional because, in certain
situations, the semantics is different depending on whether a timespan
has been specified externally or not, as we explain below.

@@

useCl past pPos (predVP (detCN (detQuant indefArt numSg) (useN (lexemeN "scandinavian_N"))) (complSlash (slashV2a (lexemeV2 "win_V2")) (detCN (detQuant indefArt numSg) (useN (lexemeN "nobel_prize_N")))))

In our semantics we deal only with present and simple past
tenses.  Indeed we find that FraCas does not exercise additional
specific tenses (if a more complicated tense is used, the additional
information is also carried by adverbs or adverbial phrases, in a more
specific way.)

@@

  ⟦``at 5 pm, s''⟧(...) = ⟦s⟧(5pm,5pm)

@@

  ⟦``in 1992, s''⟧(...) = ∃t1,t2.  [t1,t2] ⊆ 1992, ⟦s⟧(t1,t2)

@@

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  ⟦``always s''⟧(t0,t1) = ∀t0',t1'. [t0',t1'] ⊆ [t0,t1], ⟦s⟧(t0',t1')

@@

  % ⟦``every month s''⟧(t0,t1) = ∀t. [t,t+1month] ⊆ [t0,t1], ⟦s⟧(t,t+1month)
\end{enumerate}

@@

(262)
P1	Smith left after Jones left.
P2	Jones left after Anderson left.
Q 	Did Smith leave after Anderson left?
H 	Smith left after Anderson left.

@@

P-unicity : P t1 t2 → P t3 t4 → (t1 = t3) ∧ (t2 = t4)

@@

P-subsumption : [t3,t4] ⊆ [t1,t2] -> P t1 t2 -> P t3 t4

@@

P1	Smith arrived in Paris on the 5th of May, 1995.
P2	Today is the 15th of May, 1995.
P3	She is still in Paris.
Q 	Was Smith in Paris on the 7th of May, 1995?

@@

(1) Smith left at 1pm.
(2) Smith went to its appointment with the lawyer.
(3) Smith came back at 2pm
(4) Smith left at 4pm.

One would normally not say that there is contradiction. However if (1)
and (4) are presented in isolation, a contradiction can be flagged.

@@

279:
P1	Smith wrote a novel in 1991.
Q 	Did Smith write it in 1992?

Indeed, following our account, the above (contradictory) inference problem is to be
interperted as
∃x. novel(x) ∧
∃t1,t2. [t1,t2] ⊆ 1991 ∧ write(smith,x,t1,t2) ∧
∃t3,t4. [t3,t4] ⊆ 1992 ∧ write(smith,x,t3,t4)
⟶ ⊥

Thanks to unicity of action we find [t1,t2] = [t3,t4], and due to 1991
and 1992 being disjoint we obtain contradiction.

@@

However, the testsuite instructs that we should \emph{not} be able to refute
280, with the justification that ``wrote a novel'' is a repeatable
accomplishment:

280:
P1	Smith wrote a novel in 1991.
Q 	Did Smith write a novel in 1992?
H 	Smith wrote a novel in 1992.

Here our interpretation is:

∃x. novel(x) ∧
∃t1,t2. [t1,t2] ⊆ 1991 ∧ write(smith,x,t1,t2) ∧
∃y. novel(y) ∧
∃t3,t4. [t3,t4] ⊆ 1992 ∧ write(smith,x,t3,t4)
⟶ ⊥

Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t1,t2)$ and
$write(smith,y,t3,t4)$, and thus we can't deduce equality of the
interval $t1,t2$ and $t3,t4$; in turn the hypothesis can't be refuted.

@@

⟦finish to s⟧(t0,t1) = ∃(t0',t1'). t1' ∈ [t0,t1] ∧ s(t0',t1')

>>>>

\newif\ifanon\anontrue

\documentclass[a4paper,11pt]{article}
\usepackage[utf8]{inputenc}

\usepackage[hyperref]{naaclhlt2019}
\usepackage{natbib}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{todonotes}
\usepackage{enumitem}
\newlist{lingex}{enumerate}{3}
\setlist[lingex,1]{parsep=0pt,itemsep=1pt,label=(\arabic*),resume=lingexcount}
\newcommand\onelingex[1]{\begin{lingex}\item #1 \end{lingex}}

\newlist{subex}{enumerate}{3}
\setlist[subex,1]{parsep=0pt,itemsep=1pt,label=\textbf{P\arabic*}}
\newcommand\hyp{\item[\textbf{H}]}
\newcommand\fracasex[2]{\begin{lingex}\item[(#1)] \begin{subex} #2 \end{subex} \end{lingex} }
\newcommand\jp[1]{\todo[backgroundcolor=blue!10]{JP: #1}}

\newcommand\constant[1]{\mathsf{#1}}

@@

\usepackage[verbose]{newunicodechar}
\input{../PaperTools/latex/newunicodedefs}

@@

In montegovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise verb to take two
additional time parameters, one corresponding with the starting time
of the action and one corresponding to its stopping time. For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). We are thus
using a simple newtonian model of time, corresponding to layman
intuition.

@@

\newcommand\nospan{\ensuremath{─}}
\newcommand\anyspan{\ensuremath{\_}}
The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on a timespan, we will write \anyspan{} instead

@@

useCl past pPos (predVP (detCN (detQuant indefArt numSg) scandinavian\_CN) (complSlash (slashV2a win\_V2) (detCN (detQuant indefArt numSg) nobel\_prize\_CN)))

In our semantics we deal only with present and simple past
tenses. \jp{Is this enough for theory?} Indeed we find that FraCas does not exercise additional
specific tenses (if a more complicated tense is used, the additional
information is also carried by adverbs or adverbial phrases, in a more
specific way.)

@@

  \[⟦\text{at 5 pm, \(s\)}⟧(...) = ⟦s⟧(5pm,5pm)\]

@@

  \[⟦\text{in 1992, \(s\)}⟧(...) = ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)\]

@@

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \[⟦\text{always \(s\)}⟧(t_0,t_1) = ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')\]

@@

  % ⟦``every month s''⟧(t_0,t_1) = ∀t. [t,t+1month] ⊆ [t_0,t_1], ⟦s⟧(t,t+1month)
\end{enumerate}

@@

\fracasex{262}{
\item	Smith left after Jones left.
\item	Jones left after Anderson left.
\hyp 	Did Smith leave after Anderson left?
}

@@

\[\constant{unicity}_P : P t_1 t_2 → P t_3 t_4 → (t_1 = t_3) ∧ (t_2 = t_4)\]

@@

\[\constant {subsumption}_P : [t_3,t_4] ⊆ [t_1,t_2] → P t_1 t_2 → P t_3 t_4\]

@@

\fracasex{314}
{\item	Smith arrived in Paris on the 5th of May, 1995.
\item	Today is the 15th of May, 1995.
\item	She is still in Paris.
\hyp 	Was Smith in Paris on the 7th of May, 1995? 
}

@@

\begin{enumerate}[parsep=0pt,itemsep=1pt,label=(\arabic*)]
\item Smith left at 1pm.
\item Smith went to its appointment with the lawyer.
\item Smith came back at 2pm
\item Smith left at 4pm.
\end{enumerate}
One would normally not say that there is contradiction. However if (1)
and (4) are presented in isolation, a contradiction can be flagged.

@@

\fracasex{279}{
\item	Smith wrote a novel in 1991.
\hyp 	Did Smith write it in 1992?
}
Indeed, following our account, the above (contradictory) inference problem is to be
interperted as

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]

Thanks to unicity of action we find \([t_1,t_2] = [t_3,t_4]\), and due
to the years 1991 and 1992 being disjoint we obtain contradiction.

@@

However, the testsuite instructs that we should \emph{not} be able to refute
(280), with the justification that ``wrote a novel'' is a repeatable
accomplishment:

\fracasex{280}{
\item	Smith wrote a novel in 1991.
\item 	Did Smith write a novel in 1992?
\hyp 	Smith wrote a novel in 1992.
}
Here our interpretation is:

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, and thus we can't deduce equality of the
interval $t_1,t_2$ and $t_3,t_4$; in turn the hypothesis can't be refuted.

@@

\[⟦\text{finish to \(s\)}⟧(t_0,t_1) = ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')\]
